### 2013-11-18 research/paper/Theory/mathematical_model.tex
uploaded-18-11-2013

 In the QSS  model $C_y$ is determined by an interpolating polynomial based on the lift and drag data  for flow over a stationary body at various angles of attack. The order of the interpolation polynomial has varied from study to study. For  example a $7^{th}$ order polynomial was used in \cite{Parkinson1964} and $3^{rd}$ order polynomial was used in \cite{Barrero-Gil2009}. \cite{Ng2005} concluded that using a $7^{th}$ order polynomial is sufficient and a polynomial higher than that of $7^{th}$ order polynomial neither provides a significantly better result nor does it exhibit an additional amplitudes of oscillation. Thus a $7 ^{th}$ order interpolating polynomial was incorporated in this present study.

@@

\cite{Joly2012} introduced  an addition of a sinusoidal forcing function to the hydrodynamic forcing. This enables to model to provide accurate predictions even at low mass ratios where galloping excitation is suppressed or not present. In this study, the forcing due to VIV is incorporated using a sinusoidal forcing function $F_0\sin{\omega_{s}t}$ added to the RHS of the equation. $\omega_{s}$ and $F_0$ represents shedding frequency and the maximum force due to shedding respectively. Thus, the final equation is given by Eq.\eqref{final_equation_motion}.

@@

This equation could be solved by time integration methods. In  this study  `Ode 45' routine in MATLAB was used to obtain the solutions. However in certain cases (low mass ratios) `Ode 15s' routine was used when Ode 45 tend to diverge.

@@

The stationary data and the fluid-structure interaction (FSI) data were obtained using a higher order spectral element code to simulate the two-dimensional laminar flow. The Reynolds number was kept at 165 as it was pointed out by \cite{Sheard2009} and \cite{Tong2008} that the 3 dimensional transition for a square cylinder occurs at approximately Re=160. $F_0$ was kept at $0.4937$ which was obtained by scaling the value used by \cite{Joly2012} with the amplitude ratios of the lift forces obtained at the different Reynolds numbers.

@@

 FSI data were obtained for the oscillating (free-vibration) scenario. The Naiver-Stokes equations were solved using an accelerated frame of reference using the previously mentioned code. A three-step time splitting scheme together with high-order Lagrangian polynomials were used to obtain the solution. The details of the method could be found in \cite{Thompson2006,Thompson1996a}. This code was incorporated in \cite{Leontini2011,Leontini2007a}  where it was employed in a fluid-structure interaction problems.

>>>>

 In the QSS  model $C_y$ is determined by an interpolating polynomial based on the lift and drag data  for flow over a stationary body at various angles of attack. The order of the interpolation polynomial has varied from study to study. For  example a $7^{th}$ order polynomial was used in \cite{Parkinson1964} and $3^{rd}$ order polynomial was used in \cite{Barrero-Gil2009}. \cite{Ng2005} concluded that using a $7^{th}$ order polynomial is sufficient and a polynomial higher than that of $7^{th}$ order polynomial doesn't provides a significantly better result. Thus a $7 ^{th}$ order interpolating polynomial was incorporated in this present study.

@@

 It is expected that vortex shedding will be well correlated across the span and provide significant forcing at low \reynoldsnumber.\cite{Joly2012} introduced  an addition of a sinusoidal forcing function to the hydrodynamic forcing to model this. This enables to model to provide accurate predictions even at low mass ratios where galloping excitation is suppressed or not present. In this study, the forcing due to vortex shedding in low \reynoldsnumber cases is incorporated using a sinusoidal forcing function $F_0\sin{\omega_{s}t}$ added to the RHS of the equation. $\omega_{s}$ and $F_0$ represents shedding frequency and the maximum force due to shedding respectively. Thus, the final equation is given by Eq.\eqref{final_equation_motion}.

@@

This equation could be solved by time integration methods. In  this study  `Ode 45' routine in MATLAB was used to obtain the solutions. However in certain cases (low mass ratios) `Ode 15s' routine was used when Ode 45 had issues with stability due to the non-linear nature of the forcing.

@@

The stationary data and the fluid-structure interaction (FSI) data were obtained using a high order spectral element routine to simulate the two-dimensional laminar flow. The Reynolds number was kept at 165 as it was pointed out by \cite{Sheard2009} and \cite{Tong2008} that the 3 dimensional transition for a square cylinder occurs at approximately Re=160. $F_0$ was kept at $0.4937$ which was obtained by scaling the value used by \cite{Joly2012} with the amplitude ratios of the lift forces obtained at the different Reynolds numbers.

@@

 Simulations involving fluid structure interaction (FSI) were used to provide additional validation of the QSS model. The inlet was placed $20D$ while the outlet situated $60D$ away from the centroid of the body. The side boundaries were placed $20D$ away from the centroid of the body where $D$ was kept as unity throughout this study. The Naiver-Stokes equations were solved using an accelerated frame of reference using the previously mentioned code. A three-step time splitting scheme together with high-order Lagrangian polynomials were used to obtain the solution. The details of the method could be found in \cite{Thompson2006,Thompson1996a}. This code was incorporated in \cite{Leontini2011,Leontini2007a}  where it was employed in a fluid-structure interaction problems.

### 2013-11-18 research/paper/introduction/intro-1.0.tex
uploaded-18-11-2013

The search for alternate energy sources which could be categorised  under the ``green" label has become important area of research in the modern word. Solar, wind power and wave power are some of the examples of these sources.  Recently, a new branch of research has been developing to extract energy from flow induced vibrations(\cite{Bernitsas2008a-concept}). It has been hypothesized that this technique may work efficiently in areas where regular turbines cannot.

An elastically mounted slender structure such as cylinders which  are susceptible to flow-induced vibrations has the potential for energy extraction. With regards to a slender body two common types of   flow induced vibrations are Vortex Induced Vibrations (VIV) and aeroelastic galloping. Significant research has been carried out by Bernitsas and his team on extracting useful energy from VIV. Some of their significant work includes investigating the influence  of physical parameters such as mass ratio, Reynolds number, mechanical properties (\cite{Raghavan2010a} , \cite{Lee2011b}) and location (effect of the bottom boundary) (\cite{Raghavan2009}). However, the possibility of extracting energy using aeroelastic galloping has not been thoroughly investigated. Some theoretical work was carried out by (\cite{Barrero-Gil2010a}). Utilizing galloping may be a more viable method to harness energy from flow induced vibrations as it is not bounded by a narrow ``lock-in" range of reduced velocities(ratio between the freestream velocity and the product of the natural frequency of the system and the characteristic length). This study further explores the possibility of harnessing energy from flow induced vibrations using aeroelastic galloping.

Energy harvesting systems have high damping ratios because the energy generator (e.g electrical generator) puts a significant amount of damping into the system. Therefore it is crucial to investigate  the behaviour of aeroleastic galloping scenarios at high damping ratios in order to optimise the system to obtain a acceptable power output. Hence the focus of this paper is concentrated on investigating the mechanical power output of high-damped galloping systems.

According to \cite{Paidoussis2010}, \cite{Glauert1919} has provided a criterion for galloping by considering the auto-rotation of an aerofoil.  \cite{DenHartog1956}  has provided a theoretical explanation for galloping for iced electric transmission lines. A weakly non-linear theoretical aeroelastic model to predict the response of galloping was developed by \cite{Parkinson1964} based on the quasi-steady state (QSS) theory. Experimental lift and drag data on a fixed square prism at different angles of attack were used as an input for the theoretical model. It essentially used a curve fit of the transverse force to predict the galloping response. The study managed to achieve a good agreement with experimental data. The QSS model equation when solved analytically using sinusoidal solution method cannot predict cases with low mass ratios. \cite{Joly2012} have observed that finite element simulations shows a sudden change in amplitudes  below a critical values of the mass ratio. The Parkinson's equation was essentially modified to account for the vortex shedding and managed to produce the effects to the amplitude at low mass ratios.
\cite{Barrero-Gil2010a} have investigated the possibility of extracting power from vibrations caused by galloping using quasi-steady state theory. In the conclusions of that paper it was pointed out that in order to obtain a high power to area ratio, the mass-damping ($m^*\zeta$) parameter should be kept low. Another interesting conclusion was that energy conversion systems which uses galloping could operate over  a large range of flow velocities unlike VIV energy harvesting systems where the factor of energy conversion has a strong dependence on the incoming flow velocity.

>>>>

The search for alternate energy sources which could be categorised  under the ``green" label has become an important area of research in the modern word. Solar, wind power and wave power are some of the examples of these sources.  Recently, a new branch of research has been developing to extract energy from flow induced vibrations(\cite{Bernitsas2008a-concept}). It has been hypothesized that this technique may work efficiently in areas where regular turbines cannot.

An elastically mounted slender structure such as cylinders which  are susceptible to flow-induced vibrations has the potential for energy extraction. With regards to a slender bodies, two common types of   flow induced vibrations are Vortex Induced Vibrations (VIV) and aeroelastic galloping. Significant research has been carried out by Bernitsas and his team on extracting useful energy from VIV. Some of their significant work includes investigating the influence  of physical parameters such as mass ratio, Reynolds number, mechanical properties (\cite{Raghavan2010a} , \cite{Lee2011b}) and influence of proximity to solid boundary (\cite{Raghavan2009}). However, the possibility of extracting energy using aeroelastic galloping has not been thoroughly investigated. Some theoretical work was carried out by (\cite{Barrero-Gil2010a}). Utilizing galloping may be a more viable method to harness energy from flow induced vibrations as it is not bounded by a narrow ``lock-in" range of reduced velocities(\ustar). This study further explores the possibility of harnessing energy from flow induced vibrations using aeroelastic galloping.

According to \cite{Paidoussis2010}, \cite{Glauert1919} has provided a criterion for galloping by considering the auto-rotation of an aerofoil.  \cite{DenHartog1956}  has provided a theoretical explanation for galloping for iced electric transmission lines. A weakly non-linear theoretical aeroelastic model to predict the response of galloping was developed by \cite{Parkinson1964} based on the quasi-steady state (QSS) theory. Experimental lift and drag data on a fixed square prism at different angles of attack were used as an input for the theoretical model. It essentially used a curve fit of the transverse force to predict the galloping response. The study managed to achieve a good agreement with experimental data. The QSS model equation when solved analytically using sinusoidal solution method cannot predict cases with low mass ratios. \cite{Joly2012} have observed that finite element simulations shows a sudden change in amplitudes  below a critical values of the mass ratio. The Parkinson's equation when modified to account for the vortex shedding and and solved numerically, to predict the reduced amplitude at low mass ratios to the point where galloping is no longer present. \cite{Barrero-Gil2010a} have investigated the possibility of extracting power from vibrations caused by galloping using the quasi-steady state model. In the conclusions of that paper it was pointed out that in order to obtain a high power to area ratio, the mass-damping ($m^*\zeta$) parameter should be kept low. He studied the influence of the characteristics of the $C_y$ curve on maximum power output.

### 2013-11-18 research/paper/nomenclature.tex
Sketch completed..
uploaded-18-11-2013

\textbf{Nomenclature}

\begin{tabular}{ll}
$a_1,a_3,a_5,a_7$ & coefficients of the polynomial to determine $C_y$ \\ 
$C_y$ & instantaneous lift \\
$F_y$ & force due to $C_y$  \\ 
$\rho$ & fluid density  \\
$m$ & mass of the body \\
$m_a$ & added mass \\
$c$ & damping constant/damping factor \\
$k$ & spring constant \\
$U_i$ & induced velocity \\
$\theta$ & induced angle of incidence \\
$U$ & freestream velocity \\
$y,\dot{y},\ddot{y}$ & transverse displacement, velocity and acceleration   \\
$A$ & displacement amplitude\\
$\mathcal{A}$ & cross sectional area\\ 
$F_0$& force due to shedding \\
$\omega_s$& vortex shedding frequency \\
$t$ & time \\
$P_{mean}$& mean power \\
$f=\frac{1}{2\pi}\sqrt{\frac{k}{m}}$& natural frequency of the system \\
$\omega_n= 2 \pi f$& natural frequency of the system  \\
$D$ & characteristic length of the body  \\
$m^*=\frac{m}{\rho \times \text{Volume of the body}}$& mass ratio \\
$U^*=\frac{U}{f \times D}$& reduced velocity  \\
$\zeta= \frac{c}{2 m \omega_n}$& damping ratio \\
$P_t$   & power transferred to the body by the fluid \\
$P_d$& power dissipated due to mechanical damping  \\
$Re$& Reynolds number  \\
$\theta= \tan^{-1}{(\frac{\dot{y}}{U})}$& instantaneous angle \\
& \\
& \\
\end{tabular}

>>>>

\section*{Nomenclature}
%\textbf{Nomenclature}

\begin{tabular}{ll}
$a_1,a_3,a_5,a_7$ & coefficients of the polynomial to determine $C_y$ \\ 
$C_y$ & instantaneous force coefficient normal to the freesream velocity \\
$F_y$ & force due to $C_y$  \\ 
$\rho$ & fluid density  \\
$m$ & mass of the body \\
$m_a$ & added mass \\
$c$ & damping constant/damping factor \\
$k$ & spring constant \\
$U_i$ & induced velocity \\
$\theta$ & induced angle of incidence \\
$U$ & freestream velocity \\
$y,\dot{y},\ddot{y}$ & transverse displacement, velocity and acceleration   \\
$A$ & displacement amplitude\\
$\mathcal{A}$ & cross sectional area\\ 
$F_0$& force due to shedding \\
$\omega_s$& vortex shedding frequency \\
$t$ & time \\
$P_{mean}$& mean power \\
$f=\frac{1}{2\pi}\sqrt{\frac{k}{m}}$& natural frequency of the system \\
$\omega_n= 2 \pi f$& natural frequency of the system  \\
$D$ & characteristic length of the body  \\
$m^*=\frac{m}{\rho \times \text{Volume of the body}}$& mass ratio \\
$U^*=\frac{U}{f \times D}$& reduced velocity  \\
$\zeta= \frac{c}{2 m \omega_n}$& damping ratio \\
$P_t$   & power transferred to the body by the fluid \\
$P_d$& power dissipated due to mechanical damping  \\
$Re$& Reynolds number  \\
$\theta= \tan^{-1}{(\frac{\dot{y}}{U})}$& instantaneous angle \\
& \\
& \\
\end{tabular}

### 2013-11-18 research/paper/results/results-2.3.tex
uploaded-18-11-2013

The stationary $C_y$ data obtained using flow simulations (Fig.\ref{cy ploynomial}(a)) have a good agreement with data presented in \cite{Joly2012}. In comparison with the 7th order polynomial curve at \reynoldsnumber=22300 (Fig.\ref{cy ploynomial}(b)),  several differences could be observed. The peak value of $C_y$ is  significantly low at \reynoldsnumber=165 ($C_y=0.05$ at $4^0$) in comparison with \reynoldsnumber=22300 ($C_y=0.57$ at $13^0$) . The inflection point present at \reynoldsnumber=22300 could not be observed at \reynoldsnumber=165. This agrees with the findings of \cite{Luo2003}. It was concluded by \cite{Luo2003} that hysteresis occur due to the inflection point found in the $C_y$ curve. Therefore hysteresis is not expected at \reynoldsnumber=165. The range of the incident flow angles where $C_y$ remain positive is narrow at \reynoldsnumber=165 ($0^0 <\theta \leq$ $6^0$) compared to \reynoldsnumber=22300 ($0^0 <\theta \leq 15^0$). This feature is what sustains galloping and power is transferred from the fluid to the supporting structure within this range of incident angles because fluid forces are acting in the direction of travel of the oscillating cylinder. Incident angles beyond this range actually suppresses the galloping and power goes in the opposite direction. Therefore due to the overall smaller $C_y$ and narrow range of angles where $C_y$ is positive for \reynoldsnumber=165 compared to \reynoldsnumber=22300 leads one to expect a significant reduction in power at\reynoldsnumber=165 compared to \reynoldsnumber=2230.

>>>>

The stationary the force characteristic data ($C_y$) as a function of incident angle obtained using flow simulations (Fig.\ref{cy ploynomial}(a)) have a good agreement with data presented in \cite{Joly2012}. In comparison with the 7th order polynomial curve at \reynoldsnumber=22300 (Fig.\ref{cy ploynomial}(b)),  several differences could be observed. The peak value of $C_y$ is  significantly low at \reynoldsnumber=165 ($C_y=0.05$ at $4^0$) in comparison with \reynoldsnumber=22300 ($C_y=0.57$ at $13^0$) . The inflection point present at \reynoldsnumber=22300 could not be observed at \reynoldsnumber=165. This agrees with the findings of \cite{Luo2003}. It was concluded by \cite{Luo2003} that hysteresis occur due to the inflection point found in the $C_y$ curve. Therefore hysteresis is not expected at \reynoldsnumber=165. The range of the incident flow angles where $C_y$ remain positive is narrow at \reynoldsnumber=165 ($0^0 <\theta \leq$ $6^0$) compared to \reynoldsnumber=22300 ($0^0 <\theta \leq 15^0$). This feature is what sustains galloping and power is transferred from the fluid to the supporting structure within this range of incident angles because fluid forces are acting in the direction of travel of the oscillating cylinder. Incident angles beyond this range actually suppresses the galloping and power goes in the opposite direction. Therefore due to the overall smaller $C_y$ and narrow range of angles where $C_y$ is positive for \reynoldsnumber=165 compared to \reynoldsnumber=22300, it is expected that power output at \reynoldsnumber=165 is significantly lower than at \reynoldsnumber=22300.
